Model Finding for Recursive Functions in SMT

### 2016-01-26 ijcar2016/body.tex
tuned 4.1

\cvc implements a few optimizations designed to help finding finite models of
$\conv(\varphi)$.
Like other systems,
the finite model finding capability of \cvc incrementally fixes bounds on the cardinalities of uninterpreted types,
and increases these bounds until it encounters a model.
When multiple sorts are present, it uses a fairness scheme which
bounds the sum of cardinalities of all uninterpreted types~\cite{reynolds2013finite}.
For instance, if a signature has two uninterpreted types $\tau_1$ and $\tau_2$,
it will first search for models where 
the cardinality of $\tau_1$ plus the cardinality of $\tau_2$ is at most $2$,
then $3$, $4$, and so on.
To accelerate the search for models,
we use an optimization based on statically inferring \emph{monotone} types.
A monotone type is one in which models can always be extended with an additional element of that type.
Inference of monotone types has been show in previous work~\cite{claessen2011mono,korovin-2013}
to be an important component of finite model finding systems.
Moreover, it is easy to see that types $\fargtype{\con{f}}$ introduced by our translation $\conv$ are monotone.
The finite model finding capability of \cvc may take advantage of this fact by fixing the bounds for all 
monotone types simultaneously.
That is, if $\tau_1$ and $\tau_2$ are both monotone types (either introduced by our translation, or are otherwise inferred to be),
then we fix the bound for \emph{both} of these types to be $1$, then $2$, and so on.
This scheme allows the solver greater flexibility with respect to the default scheme,
and comes with no loss of generality with respect to models, since monotone sorts can always be extended to have equal cardinalities.

By default, 
\cvc uses techniques to minimize the number of literals it considers when constructing propositional satisfying assignments for formulas~\cite{relevancy2007}.
However, we have found such techniques degrade performance for finite model finding on
benchmarks having recursive functions that are defined by cases.
For this reason, we prefer disabling such techniques for benchmarks coming from our translation.

\ajr{More?}

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs heuristic methods for quantifier instantiation~\cite{MouraBjoerner07}
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered five configurations of \cvc~\cite{barrett-et-al-2011},
which we will refer to as \cvcd, \cvcf, \cvcfe, and \cvcfm.
The configuration \cvcd runs heuristic 
and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The other configurations run the finite model
finding procedure due to Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}.
The configuration \cvcfe additionally incorporates heuristic quantifier instantiation as described in Section 2.3 of~\cite{reynolds-et-al-2013},
and \cvcfm incorporates the fairness scheme for monotonic sorts as described in Section~\ref{ssec:cvc4}.

>>>>

\cvc implements a few optimizations designed to help finding finite models of
$\conv(\varphi)$.
Like other systems,
the finite model finding capability of \cvc incrementally fixes bounds on the cardinalities of uninterpreted types
and increases these bounds until it encounters a model.
When multiple types are present, it uses a fairness scheme that
bounds the sum of cardinalities of all uninterpreted types~\cite{reynolds2013finite}.
For example, if a signature has two uninterpreted types $\tau_1$ and $\tau_2$,
it will first search for models where 
$\left\vert\tau_1\right\vert + \left\vert\tau_2\right\vert$ is at most $2$,
then $3$, $4$, and so on.
To accelerate the search for models,
we implemented an optimization based on statically inferring \emph{monotonic} types.
A monotonic type is one in which models can always be extended with additional
elements of that type \cite{blanchette-krauss-2011,claessen-et-al-2011}.
%%% I find this part more confusing than helping, since we use monotonicity to merge
%%% sorts (or at least their cardinality bounds), whereas Paradox and iProver use
%%% it to separate sorts! Our use of monotonicity is more related to Nitpick's. --jb
%%%
% korovin-2013
%Inference of monotonic types has been show in previous work~
%to be an important component of finite model finding systems.
Types $\fargtype{\con{f}}$ introduced by our translation $\conv$ are monotonic,
because $\teq$ is never used directly on such types \cite{claessen-et-al-2011}.
%
%The finite model finding capability of
\cvc takes advantage of this by fixing the bounds for all 
monotonic types simultaneously.
That is, if $\tau_1$ and $\tau_2$ are inferred monotonic (whether they are
introduced by our translation or are present in the original problem),
the solver fixes the bound for \relax{both} types to be $1$, then $2$, and so on.
This scheme allows the solver greater flexibility compared with the default scheme,
and comes with no loss of generality with respect to models,
since monotonic types can always be extended to have equal cardinalities.

By default, \cvc uses techniques to minimize the number of literals it
considers when constructing propositional satisfying assignments for
formulas~\cite{relevancy2007}. However, we have found such techniques degrade
performance for finite model finding on problems having recursive functions
that are defined by cases. For this reason, we prefer disabling such
techniques for problems coming from our translation.

%%% \ajr{More?} -- Probably no space. Otherwise, you could mention zero sorts. --jb

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs heuristic methods for quantifier instantiation~\cite{MouraBjoerner07}
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered five configurations of \cvc~\cite{barrett-et-al-2011},
which we will refer to as \cvcd, \cvcf, \cvcfe, and \cvcfm.
The configuration \cvcd runs heuristic 
and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The other configurations run the finite model
finding procedure due to Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}.
The configuration \cvcfe additionally incorporates heuristic quantifier instantiation as described in Section 2.3 of~\cite{reynolds-et-al-2013},
and \cvcfm incorporates the fairness scheme for monotonic types as described in Section~\ref{ssec:cvc4}.
